Phylogenetic clustering with a Markov-modulated Poisson process

### 2015-10-19 doc/stats-seminar/presentation.tex
Updates to stats seminar talk

\section{Motivation: more informed phylogenetic clustering}

@@

\begin{frame}{Popular problems}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth, trim=0 11in 0 0, clip=true]{hughes2009molecular-f1}
        \end{column}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{lewis2008episodic-f2}
        \end{column}
    \end{columns}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \tiny
            Hughes, Gareth J., et al. ``Molecular phylodynamics of the heterosexual HIV epidemic in the United Kingdom.'' PLoS Pathog 5.9 (2009): e1000590.
        \end{column}
        \begin{column}{0.5\textwidth}
            \tiny
            Lewis, Fraser, et al. ``Episodic sexual transmission of HIV revealed by molecular phylodynamics.'' PLoS Med 5.3 (2008): e50.
        \end{column}
    \end{columns}
\end{frame}

@@

            \uncover<2->{
            \begin{tikzpicture}
                [every node/.style={circle, draw, fill, inner sep=6pt}]
                \node (a) { };
                \node (b) [below=of a] { };
                \node (c) [below right=of a] { };
                \node (d) [below left=of a] { };
                \node (e) [left=of a] { };
                \draw (a) -- (b);
                \draw (a) -- (c);
                \draw (a) -- (d);
                \draw (a) -- (e);
            \end{tikzpicture}
            }
        \end{column}
        \begin{column}{0.5\textwidth}
            \centering
            \includegraphics[height=0.8\textwidth, angle=-90]{slowtree}

@@

\section{A Markov-modulated Poisson process on phylogenies}

\begin{frame}{Modelling branching rates}
    \begin{tikzpicture}
        [every node/.style={inner sep=2pt}]
        \coordinate (a);
        \coordinate [below right=3 of a] (b);
        \draw [thick] (a) -- ++ (-0.5, 0.5) node [anchor=south east] {$\ddots$};
        \draw [thick] (a) -- ++ (-0.5, -0.4) node [inner ysep=-4pt, inner xsep=2pt, anchor=north east] {$\iddots$};
        \draw [thick] (b) -- ++ (-0.5, -0.4) node [inner ysep=-4pt, inner xsep=2pt, anchor=north east] {$\iddots$};
        \draw [thick] (b) -- ++ (0.5, -0.5) node [anchor=north west, inner ysep=-4pt] {$\ddots$};
        \draw [thick] (a) -- (b);
        \uncover<2->{
            \draw [blue, very thick] (a) -- node [auto, swap, text=blue] {slow} (b);
            \node (note1) [above right=0.5 and 6 of a, text width=4cm] {branching rates \\ along phylogeny};
        }
        \uncover<3>{
            \draw [blue] (a) -- node [auto, text=black] {
                $T \sim $ Exponential(\color{blue}$\lambda_\text{slow}$\color{black})
            } (b);
        }
        \uncover<4->{
            \draw [blue] (a) -- node [auto, text=black] {
                $T \sim $ Exponential(?)
            } (b);
        }
        \uncover<3->{
            \node (note2) [below=0.5 of note1, text width=4cm] {branching is \\ a Poisson process};
        }
        \uncover<4->{
            \node [below=0.5 of note2, text width=4cm] {rate switching is \\ a Poisson process};
            \draw [very thick, red] (b) -- node [auto] {fast} ++ (-0.5, 0.5);
        }
    \end{tikzpicture}

@@

\begin{frame}{Markov-modulated Poisson processes}
    \includegraphics[width=\textwidth]{muscariello2004mmpp}

    \includegraphics[width=\textwidth]{ching1997markov}

    \includegraphics[width=\textwidth]{lu2012markov}
\end{frame}

@@

    For a tree, we take the product of the left and right children's likelihoods.
    \begin{align*}
        L_0(T) = &\left[P_{00}(t_1)\,\lambda_0\,L_0(T_L) + P_{01}(t_1)\,\lambda_1\,L_1(T_L)\right] \\
        \times &\left[P_{00}(t_1)\,\lambda_0\,L_0(T_R) + P_{01}(t_1)\,\lambda_1\,L_1(T_R)\right].
    \end{align*}
\end{frame}

@@

    \begin{align*}
        L_0(T) = &\uncover<2->{\alert{\max}}\left(P_{00}(t_1)\,\lambda_0\,L_0(T_L) 
        \uncover<2->{,}\uncover<1>{\alert{+}}
        P_{01}(t_1)\,\lambda_1\,L_1(T_L)\right) \times \\
        &\uncover<2->{\alert{\max}}\left(P_{00}(t_1)\,\lambda_0\,L_0(T_R)
        \uncover<2->{,}\uncover<1>{\alert{+}} P_{01}(t_1)\,\lambda_1\,L_1(T_R)\right).
    \end{align*}
\end{frame}

@@

@@

\begin{frame}{Phylogenetic estimation of transmitted/founder viruses}
    \includegraphics[width=\textwidth]{mccloskey2014evaluation_f1}
\end{frame}

@@

    \vspace{1cm}
    \includegraphics[width=0.35\textwidth]{cfe}
    \includegraphics[width=0.35\textwidth]{btp}
    \includegraphics[width=0.1\textwidth]{ubc}
    \hspace{0.5cm}
    \includegraphics[width=0.1\textwidth]{ubcstats}
\end{frame}

>>>>

\section{Motivation: phylogenetic clustering}

@@

%\begin{frame}{Popular problems}
%    \begin{columns}
%        \begin{column}{0.5\textwidth}
%            \includegraphics[width=\textwidth, trim=0 11in 0 0, clip=true]{hughes2009molecular-f1}
%        \end{column}
%        \begin{column}{0.5\textwidth}
%            \includegraphics[width=\textwidth]{lewis2008episodic-f2}
%        \end{column}
%    \end{columns}
%    \begin{columns}
%        \begin{column}{0.5\textwidth}
%            \tiny
%            Hughes, Gareth J., et al. ``Molecular phylodynamics of the heterosexual HIV epidemic in the United Kingdom.'' PLoS Pathog 5.9 (2009): e1000590.
%        \end{column}
%        \begin{column}{0.5\textwidth}
%            \tiny
%            Lewis, Fraser, et al. ``Episodic sexual transmission of HIV revealed by molecular phylodynamics.'' PLoS Med 5.3 (2008): e50.
%        \end{column}
%    \end{columns}
%\end{frame}

@@

            \uncover<2->{
            \begin{tikzpicture}
                [every node/.style={circle, draw, fill, inner sep=6pt}]
                \node (a) { };
                \node (b) [below=of a] { };
                \node (c) [below right=of a] { };
                \node (d) [below left=of a] { };
                \node (e) [left=of a] { };
                \draw (a) -- (b);
                \draw (a) -- (c);
                \draw (a) -- (d);
                \draw (a) -- (e);
                \draw (b) -- (e);
                \draw (c) -- (e);
                \draw (c) -- (d);
                \draw (d) -- (e);
            \end{tikzpicture}
            }
        \end{column}
        \begin{column}{0.5\textwidth}
            \centering
            \includegraphics[height=0.8\textwidth, angle=-90]{slowtree}

@@

\section{Methods: a Markov-modulated Poisson process on phylogenies}

\begin{frame}{Modelling branching rates}
    \begin{tikzpicture}
        [every node/.style={inner sep=2pt}]
        \coordinate (a);
        \coordinate [below right=3 of a] (b);
        \draw [thick] (a) -- ++ (-0.5, 0.5) node [anchor=south east] {$\ddots$};
        \draw [thick] (a) -- ++ (-0.5, -0.4) node [inner ysep=-4pt, inner xsep=2pt, anchor=north east] {$\iddots$};
        \draw [thick] (b) -- ++ (-0.5, -0.4) node [inner ysep=-4pt, inner xsep=2pt, anchor=north east] {$\iddots$};
        \draw [thick] (b) -- ++ (0.5, -0.5) node [anchor=north west, inner ysep=-4pt] {$\ddots$};
        \draw [thick] (a) -- (b);
        \uncover<2->{
            \draw [blue, very thick] (a) -- node [auto, swap, text=blue] {slow} (b);
            \node (note1) [above right=0.5 and 6 of a, text width=4cm] {branching rates \\ along phylogeny};
        }
        \uncover<3>{
            \draw [blue] (a) -- node [auto, text=black] {
                $T \sim $ Exponential(\color{blue}$\lambda_\text{slow}$\color{black})
            } (b);
        }
        \uncover<4->{
            \draw [blue] (a) -- node [auto, text=black] {
                $T \sim $ Exponential($\cdots$)
            } (b);
        }
        \uncover<3->{
            \node (note2) [below=0.5 of note1, text width=4cm] {branching is \\ a Poisson process};
        }
        \uncover<4->{
            \node [below=0.5 of note2, text width=4cm] {rate switching is \\ a Poisson process};
            \draw [very thick, red] (b) -- node [auto] {fast} ++ (-0.5, 0.5);
        }
    \end{tikzpicture}

@@

%\begin{frame}{Markov-modulated Poisson processes}
%    \includegraphics[width=\textwidth]{muscariello2004mmpp}
%
%    \includegraphics[width=\textwidth]{ching1997markov}
%
%    \includegraphics[width=\textwidth]{lu2012markov}
%\end{frame}

@@

    For a tree, we take the product of the left and right children's likelihoods.
    \begin{align*}
        L_0(T) = &\left[P_{00}(t_1)\,\lambda_0\,L_0(T_L) + P_{01}(t_1)\,\lambda_1\,L_1(T_L)\right] \\
        \times &\left[P_{00}(t_1)\,\lambda_0\,L_0(T_R) + P_{01}(t_1)\,\lambda_1\,L_1(T_R)\right].
    \end{align*}

    \tiny
    Felsenstein, Joseph. ``Evolutionary trees from DNA sequences: a maximum
    likelihood approach.'' Journal of molecular evolution 17.6 (1981): 368-376.\par
\end{frame}

@@

    \begin{align*}
        L_0(T) = &\uncover<2->{\alert{\max}}\left(P_{00}(t_1)\,\lambda_0\,L_0(T_L) 
        \uncover<2->{,}\uncover<1>{\alert{+}}
        P_{01}(t_1)\,\lambda_1\,L_1(T_L)\right) \times \\
        &\uncover<2->{\alert{\max}}\left(P_{00}(t_1)\,\lambda_0\,L_0(T_R)
        \uncover<2->{,}\uncover<1>{\alert{+}} P_{01}(t_1)\,\lambda_1\,L_1(T_R)\right).
    \end{align*}

    \tiny
    Pupko, Tal, et al. ``A fast algorithm for joint reconstruction of ancestral
    amino acid sequences.'' Molecular Biology and Evolution 17.6 (2000):
    890-896.\par
\end{frame}

@@

\begin{frame}{Long terminal branches}
    todo
\end{frame}

\begin{frame}{Phylogenetic clusters $\neq$ contact network clusters}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{ba-net}
        \end{column}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{ba-tree}
        \end{column}
    \end{columns}
\end{frame}

@@

\begin{frame}{Phylogenetic estimation of transmitted/founder viruses}
    \begin{center}
        \includegraphics[width=0.9\textwidth]{mccloskey2014evaluation_f1}
    \end{center}
\end{frame}

@@

    \includegraphics[width=0.2\textwidth]{nserc}
    \hspace{0.2cm}
    \includegraphics[width=0.15\textwidth]{sfu}

    \includegraphics[width=0.35\textwidth]{cfe}
    \includegraphics[width=0.35\textwidth]{btp}
    \includegraphics[width=0.1\textwidth]{ubc}
    \hspace{0.5cm}
    \includegraphics[width=0.1\textwidth]{ubcstats}
\end{frame}
